An Accurate Model for SARS-CoV-2 Pooled RT-PCR Test Errors

### 2021-09-09 paper.tex
before merging uris changes

\affil[3]{The Biostatistics and Biomathematics Unit, The
Gertner Institute for Epidemiology and Health Policy Research, Sheba Medical Center, Tel Hashomer, 52621 Ramat Gan, Israel}
\affil[*]{Equal contribution}
\date{}

@@

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the application of techniques to increase the throughput
of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman,
  MatrixPooling, LionDorfman}. These techniques, often termed
\emph{pooling}, or group testing, have originated in the 1940's in the
seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}. Other commonly used techniques include
recursive pooling, wherein... Matrix pooling.. many other complex
techniques..

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed
that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool},
as well as in more specific, COVID-19 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge this
commonly made assumption and show how using a more accurate
probabilistic model affects the estimation of false-negative
rates for Dorfman pooling.

@@

\section*{Discussion}
In this study, we developed a more realistic and novel probabilistic
model for the outcomes of pooled PCR tests, parametrized for
SARS-CoV-2. Contrary to the common assumption, we assume, based on
data (Table~\ref{table}), that multiple infected individuals increase
the likelihood of a positive pooled test result. A direct consequence
of our model is that low values of infection
prevalence increase the false-negative rates of Dorfman pooling. Importantly, our results remain qualitatively similar under varying parameter values,
in the observed ranges \cite{KitComparison,EstimatingRatesKucrika,
  EstimatingRatesLourenco, InterpretingCOVID19Test} (Figure
\ref{fig1}). Therefore, our results give rise to a conflict: low infection
prevalence leads to high efficiency of Dorfman pooling
\cite{DorfmanYuvalDor}, while also increasing false-negative rates.

@@

Furthermore, an especially important
consideration in designing pooling schemes is explicitly taking the
intrinsic PCR error rates into account. As we have demonstrated above, analyzing the expected error rates of pooling schemes under a realistic probabilistic error model is readily achievable. Designing pooling schemes which optimize testing while accounting for error rates is, however, far from trivial. Nonetheless, various algorithms explicitly incorporate error rates into their pooling optimization processes (REFS), including a recent study by our group which employed the herein described error model \cite{DOPE}.

>>>>

\affil[3]{The Biostatistics and Biomathematics Unit, The Gertner
  Institute for Epidemiology and Health Policy Research, Sheba Medical
  Center, Tel Hashomer, 52621 Ramat Gan, Israel}

\affil[*]{Equal contribution}

\date{}

@@

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the application of techniques to increase the throughput
of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman,
  MatrixPooling, LionDorfman, CherifReview}. These techniques, often
termed \emph{pooling}, or group testing, have originated in the 1940's
in the seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroguhly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}. Other commonly used techniques include
\emph{recursive} pooling \cite{Kim,RecursiveSevenFold}, wherein if the
first pooled test is positive, the pool is split into two and the
process repeats. Otherwise, all pool members are declared negative.
In \emph{Matrix} pooling \cite{MatrixPooling} a population of size
$N=mn$ is arranged in a matrix of dimensions $m\times n$ matrix. Each
row and column are then pooled and individuals in the intersection of
positive rows and columns are tested separately.

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID, CherifReview}. Consequently,
studies assumed that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool,
  CherifReview}, as well as in more specific, COVID-19 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge this
commonly made assumption and show how using a more accurate
probabilistic model affects the estimation of false-negative rates for
Dorfman pooling.

@@

\section*{Discussion}
In this study, we developed a more realistic and novel probabilistic
model for the outcomes of pooled PCR tests, parametrized for
SARS-CoV-2. Contrary to the common assumption, we assume, based on
data (Table~\ref{table}), that multiple infected individuals increase
the likelihood of a positive pooled test result. A direct consequence
of our model is that false-negative rates depend on infection
prevalence. A further consequence is that low values of infection
prevalence increase the false-negative rates of Dorfman pooling. These
results remain qualitatively similar under varying parameter values,
in the observed ranges \cite{KitComparison,EstimatingRatesKucrika,
  EstimatingRatesLourenco, InterpretingCOVID19Test} (Figure
\ref{fig1}). Our results give rise to a conflict: low infection
prevalence leads to high efficiency of Dorfman pooling
\cite{DorfmanYuvalDor}, while also increasing false-negative rates.

@@

Thus, improving pooling schemes is imperative. An especially important
consideration in designing such schemes is explicitly taking the
intrinsic PCR error rates into account. As we have demonstrated above,
analyzing expected error rates of a given scheme, given some error
model is relatively straightforward. Designing a pooling scheme
tailored for a specific error model is, however, far from trivial. In
a recent study \cite{DOPE}, we have developed a novel pooling scheme
that takes advantage of this specific error model. Our new pooling
scheme shows great promise in terms of both accuracy, as well as
efficiency, although it has not yet been implemented in real-world
settings.
